Ion pairing controls rheological properties of ``crocodile-line'' polyelectrolyte hydrogels

### 2016-01-29 PALSE/ion_pairing_paper.tex
Corrections Seb.

\begin{abstract}
We demonstrated recently that polyelectrolytes with cationic moieties along the chain and a single anionic head are able to form physical hydrogels due to the reversible nature of the head-to-body ionic bond. Here we generate a variety of such polyelectrolytes with various cationic moieties and counter-ions combinations starting from a common polymeric platform. We show that the rheological properties (shear modulus, critical strain) of the final hydrogels vary by three orders of magnitude depending on the cation/pair. Our data are well described by a theoretical model involving a supramolecular crocodile line between cross-links, the length of the line being set by the amount of charge dissociation.
\end{abstract}

\section{Introduction}
Since the pioneering work of \citeauthor{Wichterle1960} which established their relevance in a biomedical context, polymer based hydrogels have never ceased to be a very active field of research\cite{Calo2015,Hoffman2001}. They have recently experienced a burst of interest among the biomedical community as controlled drug delivery cargos or scaffolds for wound healing or tissues repair\cite{Vashist2014,Ratner2004}. Polyelectrolytes are being extensively used in this particular context\cite{Rosso2003,BinImran2014}. Their gel formation mechanism often involves reversible-by-nature electrostatic interactions, which can be used for instance to trigger ``smart'' release of bioactive substances\cite{Khare1993,Lockwood2007}. Besides, poly(cationic) gel have been reported to combine scaffold properties for cell’s adherence and growth with antimicrobial activity, and are therefore increasingly used in clinical applications\cite{Hoffman2001,Landers2002}. In this framework, injectability is a key feature, as it provides an easier way to gel delivery in vivo\cite{Tibbitt2016}; thus, reversible shear thinning biocompatible hydrogels are a particularly sought after class of materials\cite{Yu2008}.

We have recently reported on a highly reversible poly(cation) based hydrogel\cite{Srour2014}, which formation relies on an new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate terminated ATRP initiator, eventually affording an anion terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.

In the present report, we investigate the role of the dissociation rate of the poly(cationic) body on the mechanical/rheological properties of the hydrogel. In a first part we describe how we take advantage of our well-controlled synthesis to play systematically with the nature (hardness/softness) of the cationic repeating unit and the ion-paring strength of the associated counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}). In a second part, we show the qualitative and quantitative variations of mechanical properties of the aqueous dispersion obtained from these well-characterized polymers. In a third part, we develop a microscopic model based on the idea that at low dissociation rates, cross-links are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular crocodile line. Finally, we discuss our results and possible applications of such smart hydrogels.

@@

We synthesise the phosphonate terminated polymer \ce{PBr} according to our previously reported methodology\cite{Srour2014,Appukuttan2012}. We obtain a well-controlled linear polymer ($Mw= \SI{5614}{\dalton}$, $Mw/Mn = 1.08$; S2, S3, Figure S1), of polymerisation index $N_0=70$, which is the common basis to all studied polymers. We achieve all subsequent transformations of the polymer structures using the same initial batch of material. Nucleophilic addition of N-Methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF affords the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Scheme~\ref{sch:synthesis}).

\begin{scheme}
\includegraphics{synthesis.pdf}
\caption{Synthesis of \ce{PIm+Br-} and \ce{PPyr+Br-} and their intermediates \ce{POH} and \ce{PBr}. $N_0=70$}
\label{sch:synthesis}
\end{scheme}

We performed anionic metathesis by pouring an aqueous solution of \ce{PIm+Br-} or \ce{PPyr+Br-} into a saturated aqueous solution of the different sodium halides (\ce{NaF}, \ce{NaCl}, \ce{NaI}) (Scheme~\ref{sch:metathesis}). In the cases of \ce{PIm+F-}, \ce{PIm+Cl-} , \ce{PPyr+F-} and \ce{PPyr+Cl-}, we obtain a turbid suspension immediately after addition. After extensive dialysis of the resulting mixture against deionized water and lyophilization we recover in high yields the different \ce{PIm+X-} and \ce{PPyr+X-} (where \ce{X}=\ce{F}, \ce{Cl}, \ce{I}). In order to bring evidence for the efficiency of the ionic metathesis, we submit the resulting materials to negative mode high resolution mass spectrometry (HRMS, S3). We unambiguously assess displacement of bromide counterions by the full disappearance of the diisotopic mass peak (89/\SI{91}{\dalton}, Figures S8-9). We further confirm substitution by iodide by the concomitant apparition of a characteristic monoisotopic peak (\SI{121}{\dalton}, Figures S6-7), while fluoride and chloride anions signals, which are below the detection limits of the spectrometer, could not be observed in the corresponding polymers (see supporting information).

@@

In ionic solids, melting points are inherently connected to the reticular enthalpy of the crystal lattice, hence primarily to ion pairing strength\cite{Sherman1932}. This last parameter is directly related to the diffuse and hindered nature of the interacting charges, bulkier and softer ions providing in the most general case weaker interactions\cite{Krossing2006}. For typical ``ionic liquids'' such as imidazolium or pyrollidinium halides, melting points will thus gradually increase as the size of the halide counterion decreases, ie $\ce{I-}<\ce{Br-}<\ce{Cl-}<\ce{F-}$\cite{Dean2010}. Although analysis is less straightforward in the case of amorphous polymer electrolytes, ionic substitution is however expected to have an impact on the phase transition temperatures of the material\cite{Orler1994}.

@@

In the case polyelectrolytes with fluoride and chloride counterion, we observe no exo or endothermic transition below \SI{200}{\celsius}. By contrast, we observe a broad endothermic peak when iodide and bromide are used as counterions. Quite remarkably,we found a similar peak temperature (84-\SI{85}{\celsius}) for \ce{PIm+I-} and \ce{PPyr+I-}, see Figure~\ref{fig:dsc}. When bromide counterions are present, we observe a significant increase of the peak temperature (114-\SI{115}{\celsius}) but again, with a similar value between \ce{PIm+Br-} and \ce{PPyr+Br-} as shown in Figure~\ref{fig:dsc}. The overall shape and position of the DSC peaks is very reminiscent of previously reported data for various naturally occuring or synthetic poly(electrolytes)\cite{Li2005,Sarmento2006,Ostrowska-Czubenko2009a,Moin2015}. It is generally attributed to desorption of weakly bound water from the polymer network. Its presence in the polymer samples where bromide and iodide are used as couterions suggests that adsorbed water could play a role in solvating the dissociated ion pairs in the latter; conversely, its absence for chloride and fluoride counterions suggests that dissociation is not effective because of stronger ion pair interaction.

\begin{figure}
\includegraphics{dsc.pdf}
\caption{DSC measurements of polymer samples showing the presence of a endothermic peak. Curves are shifted vertically by arbitrary amounts for clarity. Vertical lines indicate the peak temperatures at \SI{84}{\celsius} for the two iodides in black and \SI{114}{\celsius} for the two bromides in gray.}
\label{fig:dsc}
\end{figure}

\subsection{Water swelling, ion-pairing and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer compositions. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

The above dependence of gel formability on the nature of the counterions could be rationalized on the basis of ion-pairing processes. According to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected between a ``softer'' cationic moiety and a ``harder'' halide counterions\cite{Goossens2009}. Therefore charge dissociation in water and thus solubility of the polymer should be easier with larger halide and pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge. That is why we are able to dissolve the five polymers that have the weakest ion pairing, but neither the two fluorides nor \ce{PIm+Cl-} where counterions are expected to be strongly bounded to the chain.

A fortiori, the insolubility in absence of charge dissociation indicates that the chain is in poor solvent. To estimate $\Theta$ temperature, we held the two fluoride polymers in water at boiling temperature overnight without any dissolution, implying $\Theta>\SI{100}{\celsius}$. Nonetheless, due to charge dissociation, five of our polyelectrolytes are soluble at \SI{25}{\celsius}.

@@

In Figure~\ref{fig:strain}a we compare the rheological responses of Poly(pyrrolidinium) based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same mechanical behavior in the linear regime but \ce{PPyr+I-} yields at slightly higher strains ($\gamma_c=4\%$ instead of $3\%$). By contrast, the gel with chloride counterions is weaker by more than one order of magnitude. In other words, a ``soft'' halide counterion, efficiently dissociated, is correlated to a harder gel.

@@

We have previously demonstrated that, due to the opposite charge of the phosphonate head, head-to-body ionic bonds are possible~\cite{Srour2014}. If at least two foreign heads attach to the same body, we obtain an effective cross-linking point. When the probability of such a configuration is non-zero, we obtain a physically cross-linked gel.

Here we argue that if every chain has a single foreign head attached to its body, every polymer is linked to two others in a crocodile line. We neglect the possibility of the two charges of the phosphonate head to attach to two separate chains. In addition, if we suppose that in order to minimize the inter-chain repulsion between charged cationic groups, heads are preferentially attached to the tail of their neighbour, we obtain a linear chain of effective polymerisation index $n N_0$ where $n$ is the number of polymers in the crocodile line.

We thus have two limiting cases: (i) every chain has at least two heads attached and we have roughly $N_0$ monomers between cross-linking points; (ii) every chain has a single head attached and we have a single supramolecular chain in the system. The former case should be observed at high charge dissociation when many moieties are ionised along the chain providing many sites for a phosphonate head to bound to. The latter case should be observed at negligible charge dissociation. In between we should observe cross-linking points separated by crocodile lines of $n N_0$ monomers as sketched in Figure~\ref{fig:dissociation}a.

Let us now discuss the rheological measurements and link them to microstructure. We reduce our measurements to two main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, and (ii) the critical strain $\gamma_c$ corresponding to the maximum of $G"$, the end of the linear regime and the first broken cross-links.

@@

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 790 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This is coherent with a higher probability of attaching two or more heads on a highly dissociated body, i.e. a large cross-linking ratio.

\begin{table}
\begin{tabular}{l|SS|SSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.05&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	1.1\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	45	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	7.\\
\ce{PIm+I-}	& 78	&	0.16	&	113	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	9.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	790	&	64	&	160.	&	3.4\cellcolor{gray!25}\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates incoherent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

Considering that each crocodile chain is an entropic spring in its linear domain, the mechanical energy stored by a single crocodile chain at a strain $\gamma$ is $E = \gamma^2 k_\mathrm{B}T$. If we suppose that the ionic bonds breaks before the complete extension of the crocodile chains, we evaluate the head-to-body bonding energy to
\begin{equation}
E_c = \gamma_c^2 k_\mathrm{B}T.
\label{eq:Ec}
\end{equation}

For \ce{PIm+Br-} we measure $\gamma_c = 800\%$ and thus $E_c \approx 65 k_\mathrm{B}T$ per head-to-body bond or \SI{160}{\kilo\joule\per\mol}. This value is much higher than the $E_i\approx\SI{5}{\kilo\joule\per\mol}$ typical of ionic bonds in water\cite{Schneider1992}, even when adding up the two bonds of the bivalent phosphonate head. However the surrounding medium is not water but the hydrophobic chains. Such an apolar microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly\cite{Rehm2010}. The local dielectric constant can easily be 16 times lower than water, justifying the large bonding energy we measure.

However for all other samples the critical strain is between 2\% and 20\%, leading to incoherent values for the bonding energies between $10^{-3}k_\mathrm{B}T$ and $10^{-1}k_\mathrm{B}T$ per bond, see Table~\ref{tab:results}. Therefore, the hypothesis of a linear spring fails for these four samples, meaning that the crocodile chain is fully extended at $\gamma_c$.

@@

The spatial conformation of a polyelectrolyte chain in semi-dilute, low salt solution is a persistent random walk where the persistence length is the screening length $r_\mathrm{src}$ due to the dissociated counter ions\cite{Rubinstein1996}. In bad solvent, the persistent random walk is made of electrostatic blobs of diameter $D$ containing $g_e$ monomers, see Figure~\ref{fig:dissociation}a:
\begin{align}
g_e &= \frac{A^2}{u}\tau\label{eq:ge},\\
D &= b \left(\frac{A^2}{u}\right)^{1/3}\label{eq:D}.
\end{align}
Here $A$ is the number of monomers between uncondensed charges, $\tau = 1 - T/\Theta$ the reduced temperature and $u = \ell_\mathrm{B}/b \approx 2.7$ is the ratio of Bjerrum length $\ell_\mathrm{B}$ and monomer size $b\approx \SI{0.26}{\nano\metre}$.

@@

The only unknown here is the reduced temperature $\tau$, possibly different between poly(pyrrolidinium) based and poly(immidazolium) based polymers. We know that the undissociated chain has a $\Theta$ temperature higher than \SI{100}{\celsius}, thus $0.2<\tau<1$. Setting $\tau\approx 0.235$, corresponding to $\Theta\approx\SI{116}{\celsius}$, yields $A=1$ for both \ce{PPyr+I-} and \ce{PPyr+Br-}, indicating full dissociation of the charges. By contrast \ce{PPyr+Cl-} has a dissociated charge every 7 monomers. Supposing the same $\tau$, \ce{PIm+I-} has a dissociated charge every 10 monomers, see Table~\ref{tab:results}.

@@

Finally we cross check our model by linking the charge dissociation and and the amount of physical cross-link by an independent argument. If we assume that the probability for a charge to dissociate is independent of the number of charges already present on the polymer, we can describe the probability of having $k$ charge dissociated on a polymer by a Poisson distribution of mean $\lambda = N_0/A$. Having 2 charges dissociated on a chain implies a cross-link. Therefore, we the cross-link probability $1/n$ is proportional to $\lambda^2$, yielding $n \sim A^2$.

As shown in Figure~\ref{fig:dissociation}b this relation describes very well our data. We can extrapolate this law for the case of \ce{PIm+Br-} and obtain $A = 28$. Inverting equation~(\ref{eq:A}) yields $\gamma_c = 860\%$, indeed higher than the one observed (804\%). This confirms that in the \ce{PIm+Br-} strain sweep we were actually probing the strength of the ionic bonds.

>>>>

\begin{abstract}
We demonstrated recently that polyelectrolytes with cationic moieties along the chain and a single anionic head are able to form physical hydrogels due to the reversible nature of the head-to-body ionic bond. Here we generate a variety of such polyelectrolytes with various cationic moieties and counter-ions combinations starting from a common polymeric platform. We show that the rheological properties (shear modulus, critical strain) of the final hydrogels vary by three orders of magnitude depending on the cation/anion pair. Our data are well described by a theoretical model involving a supramolecular crocodile line between cross-links, the length of the line being set by the amount of charge dissociation.
\end{abstract}

\section{Introduction}
Since the pioneering work of \citet{Wichterle1960} which established their relevance in a biomedical context, polymer-based hydrogels have never ceased to be a very active field of research\cite{Calo2015,Hoffman2001}. They have recently experienced a burst of interest among the biomedical community as controlled drug delivery cargos or scaffolds for wound healing or tissue repair\cite{Vashist2014,Ratner2004}. Polyelectrolytes are being extensively put to use in this particular context\cite{Rosso2003,BinImran2014}. Their gel formation mechanism often involves reversible-by-nature electrostatic interactions, which can be used for instance to trigger ``smart'' release of bioactive substances\cite{Khare1993,Lockwood2007}. Besides, poly(cationic) gels have been reported to combine scaffold properties for cell’s adherence and growth with antimicrobial activity, and are therefore increasingly used in clinical applications\cite{Hoffman2001,Landers2002}. In this framework, injectability is a key feature, as it provides an easier way to gel delivery in vivo\cite{Tibbitt2016}; thus, reversible shear thinning biocompatible hydrogels are a particularly sought after class of materials\cite{Yu2008}.

We have recently reported on a highly reversible poly(cation) based hydrogel\cite{Srour2014}, which formation relies on an new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate-terminated ATRP initiator, eventually affording an anion-terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.

In the present report, we investigate the role of the dissociation rate of the poly(cationic) body on the mechanical properties of the hydrogel. In a first part we describe how we take advantage of our well-controlled synthesis to play systematically with the nature (hardness/softness) of the cationic repeating unit and the ion-paring strength of the associated counter-ion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}). In a second part, we show the qualitative and quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a third part, we develop a microscopic model based on the idea that at low dissociation rates, cross-links are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular crocodile line. %Finally, we discuss our results and possible applications of such smart hydrogels.

@@

We synthesise the phosphonate terminated polymer hereafter reffered to as \ce{PBr} according to our previously reported methodology\cite{Srour2014,Appukuttan2012}. We obtain a well-controlled linear polymer ($Mw= \SI{5614}{\dalton}$, $Mw/Mn = 1.08$; see Supplementary Materials S2, S3, Figure S1), of polymerisation index $N_0=70$, which is the common basis to all studied polymers. We achieve all subsequent transformations of the polymer structures using the same initial batch of material. Nucleophilic addition of N-methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF affords the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Scheme~\ref{sch:synthesis}).

\begin{scheme}
\includegraphics{synthesis.pdf}
\caption{Synthesis of \ce{PIm+Br-} and \ce{PPyr+Br-} and their intermediates \ce{POH} and \ce{PBr} with $N_0=70$.}
\label{sch:synthesis}
\end{scheme}

We performed anionic metathesis by pouring an aqueous solution of \ce{PIm+Br-} or \ce{PPyr+Br-} into a saturated aqueous solution of the different sodium halides (\ce{NaF}, \ce{NaCl}, \ce{NaI}) (Scheme~\ref{sch:metathesis}). In the cases of \ce{PIm+F-}, \ce{PIm+Cl-} , \ce{PPyr+F-} and \ce{PPyr+Cl-}, we obtain a turbid suspension immediately after addition. After extensive dialysis of the resulting mixture against deionized water and lyophilization we recover in high yields the different \ce{PIm+X-} and \ce{PPyr+X-} (where \ce{X}=\ce{F}, \ce{Cl}, \ce{I}). In order to bring evidence for the efficiency of the ionic metathesis, we submit the resulting materials to negative-mode high-resolution mass spectrometry (HRMS, see Supplementary Materials S3). We unambiguously assess displacement of bromide counterions by the full disappearance of the diisotopic mass peak (89/\SI{91}{\dalton}, Figures S8-9). We further confirm substitution by iodide by the concomitant apparition of a characteristic monoisotopic peak (\SI{121}{\dalton}, Supplementary Figures S6-7), while fluoride and chloride anions signals, which are below the detection limits of the spectrometer, could not be observed in the corresponding polymers (see supporting information XXX S? XXX).

@@

In ionic solids, melting points are inherently connected to the reticular enthalpy of the crystal lattice, hence primarily to ion pairing strength\cite{Sherman1932}. This last parameter is directly related to the diffuse and hindered nature of the interacting charges, bulkier and softer ions providing in the most general case weaker interactions\cite{Krossing2006}. For typical ``ionic liquids'' such as imidazolium or pyrrolidinium halides, melting points will thus gradually increase as the size of the halide counterion decreases, ie $\ce{I-}<\ce{Br-}<\ce{Cl-}<\ce{F-}$\cite{Dean2010}. Although analysis is less straightforward in the case of amorphous polymer electrolytes, ionic substitution is however expected to have an impact on the phase transition temperatures of the material\cite{Orler1994}.

@@

In the case polyelectrolytes with fluoride and chloride counterion, we observe no exo- or endothermic transition below \SI{200}{\celsius}. By contrast, we observe a broad endothermic peak when iodide and bromide are used as counterions. Quite remarkably, we found a similar peak temperature (84-\SI{85}{\celsius}) for \ce{PIm+I-} and \ce{PPyr+I-}, see Figure~\ref{fig:dsc}. When bromide counterions are present, we observe a significant increase of the peak temperature (114-\SI{115}{\celsius}) but again, with a similar value between \ce{PIm+Br-} and \ce{PPyr+Br-} as shown in Figure~\ref{fig:dsc}. The overall shape and position of the DSC peaks is very reminiscent of previously reported data for various naturally occuring or synthetic poly(electrolytes)\cite{Li2005,Sarmento2006,Ostrowska-Czubenko2009a,Moin2015}. It is generally attributed to desorption of weakly bound water from the polymer network. Its presence in the polymer samples where bromide and iodide are used as couterions suggests that adsorbed water could play a role in solvating the dissociated ion pairs in the latter. Conversely, its absence for chloride and fluoride counterions suggests that dissociation is not effective because of stronger ion pair interaction.

\begin{figure}
\includegraphics{dsc.pdf}
\caption{DSC measurements of polymer samples showing the presence of an endothermic peak. Curves are shifted vertically by arbitrary amounts for clarity. Vertical lines indicate the peak temperatures at \SI{84}{\celsius} for the two iodides in black and \SI{114}{\celsius} for the two bromides in gray.}
\label{fig:dsc}
\end{figure}

\subsection{Water swelling, ion-pairing and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer composition. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

The above dependence of gel formability on the nature of the counterions could be rationalized on the basis of ion-pairing processes. According to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected between a ``softer'' cationic moiety and a ``harder'' halide counterions\cite{Goossens2009}. Therefore charge dissociation in water and thus solubility of the polymer should be easier with larger halide and pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge. That is why we are able to dissolve the five polymers that have the weakest ion pairing, but neither the two fluorides nor \ce{PIm+Cl-} where counterions are expected to be strongly bound to the chain.

A fortiori, the insolubility in absence of charge dissociation indicates that the chain is in poor solvent. To estimate the $\Theta$-temperature, we held the two fluoride polymers in water at boiling temperature overnight without any dissolution, implying that $\Theta>\SI{100}{\celsius}$. Nonetheless, due to charge dissociation, five of our polyelectrolytes are soluble at \SI{25}{\celsius}.

@@

In Figure~\ref{fig:strain}a we compare the rheological responses of Poly(pyrrolidinium)-based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same mechanical behavior in the linear regime but \ce{PPyr+I-} yields at slightly higher strains ($\gamma_c=4\%$ instead of $3\%$). By contrast, the gel with chloride counterions is weaker by more than one order of magnitude. In other words, a ``soft'' halide counterion, efficiently dissociated, is correlated to a harder gel.

@@

We have previously demonstrated that, due to the opposite charges of the phosphonate head, head-to-body ionic bonds are possible~\cite{Srour2014}. On the one hand, if at least two foreign heads attach to the same body, we obtain an effective cross-link point. When the probability of such a configuration is non-zero, we obtain a physically cross-linked gel.

On the other hand, if every chain has a single foreign head attached to its body, every polymer is linked to two others in a crocodile line\bibnote{We neglect the possibility of the two charges of the phosphonate head to attach to two separate chains}. In addition, if we suppose that in order to minimize the inter-chain repulsion between charged cationic groups, heads are preferentially attached to the tail of their neighbour, we obtain a linear chain of effective polymerisation index $n N_0$ where $n$ is the number of polymers in the crocodile line.

We thus have two limiting cases: (i) every chain has at least two heads attached and we have roughly $N_0$ monomers between cross-link points; (ii) every chain has a single head attached and we have a single supramolecular chain in the system. The former case should be observed at high charge dissociation when many moieties are ionised along the chain providing many sites for a phosphonate head to bound to. The latter case should be observed at negligible charge dissociation. In between these two limiting cases, we should observe cross-link points separated by crocodile lines of $n N_0$ monomers as sketched in Figure~\ref{fig:dissociation}a.

Let us now discuss the rheological measurements and link them to the hydrogel microstructure. We reduce our measurements to two main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, and (ii) the critical strain $\gamma_c$ corresponding to the maximum of $G"$, that we consider as representative of the end of the linear regime and of the first broken cross-links. XXX Ref XXX

@@

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 790 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This is consistent with a higher probability of attaching two or more heads on a highly dissociated body, i.e. a large cross-link ratio.

\begin{table}
\begin{tabular}{l|SS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$} & {$\sqrt{n}$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	1.	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.05&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	1.1	&	1.\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	45	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	7.	&	6.7\\
\ce{PIm+I-}	& 78	&	0.16	&	113	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	9.6	&	10.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	790	&	64	&	160.	&	3.4\cellcolor{gray!25}	&	28\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates inconsistent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

Considering that each crocodile chain is an entropic spring in its linear domain, the mechanical energy stored by a single crocodile chain under a strain $\gamma$ is $E = \gamma^2 k_\mathrm{B}T$. If we suppose that ionic bonds break before the complete extension of the crocodile chains, we evaluate the head-to-body bonding energy to
\begin{equation}
E_c = \gamma_c^2 k_\mathrm{B}T.
\label{eq:Ec}
\end{equation}

For \ce{PIm+Br-} we measure $\gamma_c = 800\%$ and thus $E_c \approx 65 k_\mathrm{B}T$ per head-to-body bond or \SI{160}{\kilo\joule\per\mol}. This value is much higher than the $E_i\approx\SI{5}{\kilo\joule\per\mol}$ typical of ionic bonds in water\cite{Schneider1992}, even when adding up the two bonds of the bivalent phosphonate head\bibnote{Since the two minus charges are born by the same atom, chelation effect may enhance bounding energy, but not by an order of magnitude}. However, we have to remember that the chain is in bad solvent, thus forming locally compact blobs almost devoid of water. Therefore, the medium surrounding the ionic bond is not water but the hydrophobic chains. Such an apolar microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly\cite{Rehm2010}. Quantitatively, the bonding energy of an ionic bond is inversely proportional to the relative dielectric constant $\epsilon_r$, therefore we need to suppose a local $\epsilon_r$ around 5 (instead of 80 in water) to recover the large bonding energy we measure.

However for all other samples the critical strain is between 2\% and 20\%, leading to inconsistent values for the bonding energies ranging between $10^{-3}k_\mathrm{B}T$ and $10^{-1}k_\mathrm{B}T$ per bond, see Table~\ref{tab:results}. Therefore, the end of the linear regime cannot correspond to bond breaking but rather to the point where extended chains become nonlinear.

@@

The spatial conformation of a polyelectrolyte chain in semi-dilute, low salt solution is a persistent random walk where the persistence length is the screening length $r_\mathrm{src}$ due to the dissociated counter ions\cite{Rubinstein1996}. In bad solvent, the persistent random walk is made of electrostatic blobs of diameter $D$ containing $g_e$ monomers, see Figure~\ref{fig:dissociation}a:
\begin{align}
g_e &= \frac{A^2}{u}\tau\label{eq:ge},\\
D &= b \left(\frac{A^2}{u}\right)^{1/3}\label{eq:D}.
\end{align}
Here $A$ is the number of monomers between dissociated ion pairs, $\tau = 1 - T/\Theta$ the reduced temperature and $u = \ell_\mathrm{B}/b \approx 2.7$ is the ratio of the Bjerrum length in water $\ell_\mathrm{B}\approx \SI{0.7}{\nano\metre}$ and monomer size $b\approx \SI{0.26}{\nano\metre}$.

@@

The only unknown here is the reduced temperature $\tau$, possibly different between poly(pyrrolidinium) based and poly(immidazolium) based polymers. We know that the undissociated chain has a $\Theta$ temperature higher than \SI{100}{\celsius}, thus $0.2<\tau<1$. In addition, we must have $A\geq 1$, which implies through (Eq.~\ref{eq:A}) $\tau\leq 0.235$, corresponding to $\Theta\approx\SI{116}{\celsius}$. This limit value yields $A=1$ for both \ce{PPyr+I-} and \ce{PPyr+Br-}, indicating full dissociation of the cationic moieties. By contrast \ce{PPyr+Cl-} has a charge every 7 monomers. Supposing the same $\tau$, \ce{PIm+I-} has a dissociated ion pair every 10 monomers, see Table~\ref{tab:results}. Of course analysing \ce{PIm+Br-} based on the value of $\gamma_c$ gives an inconsistent value $A=3.4$ indicating that the hydrogel doesn't belong to the same regime.

@@

Finally we cross-check our model by linking the ion pair dissociation and the amount of physical cross-links by an independent argument. If we assume that the probability for an ion pair to dissociate is independent of the number of dissociated pairs already present on the polymer, we can describe the probability of having $k$ dissociated pairs on a polymer by a Poisson distribution of mean $\lambda = N_0/A$. Having 2 charges dissociated on a chain implies a cross-link. Therefore, the cross-link probability $1/n$ is proportional to $\lambda^2$, yielding $n \sim A^2$.

As shown in Figure~\ref{fig:dissociation}b and the last two columns of Table~\ref{tab:results} this relation describes very well our data. We can extrapolate this law for the case of \ce{PIm+Br-} and obtain $A = 28$. Inverting (Eq.~\ref{eq:A}) yields $R_0/R - 1 = 860\%$, indeed higher than the observed $\gamma_c\approx 804\%$. This confirms that in the \ce{PIm+Br-} strain sweep $\gamma_c$ actually results from the strength of the ionic bonds.
